Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2025-07-22 revision/response.qmd

# Reviewer 1

## General comments

1. The manuscript is being considered for the special issue on Statistical Aspects of Trustworthy Machine Learning. However, the connection to the theme of the special issue feels somewhat underdeveloped. It would be helpful for the authors to clarify which aspect(s) of trustworthiness are being addressed. While the work appears to relate to robustness and reproducibility, these connections should be more clearly articulated and substantiated in the manuscript.

**\textcolor{purple}{We now discuss the connection between our work and the trustworthiness of machine learning in the third paragraph of the Introduction. Specifically, our work address the breakdown in trust caused by  violations of algorithmic assumptions -- corresponding to the third step in Broderick et al. (2023). In modern data analysis, where data are often not collected directly by the analysts, assumptions about the data can be unknowingly violated, leading to unexpected outcomes. Our approach advocates for making these assumptions -- often simple or taken for granted -- explicit, in order to more efficiently diagnose unexpected outcomes.}**

@@

2. The proposed procedure appears to assume that it is possible to simulate datasets similar to the observed data (e.g., as illustrated in Figure 2). However, this assumption may not be realistic in many practical scenarios. Simulating data requires a data-generating model, but commonly used statistical methods such as generalized estimating equations (GEE) or the generalized method of moments (GMM) do not fully specify the underlying distribution or likelihood. Even when the mean structure is specified, there may be infinitely many likelihood models consistent with it. Simulating from only one such model risks overlooking a broad class of plausible alternatives. How do the authors propose to handle such cases? More broadly, generating synthetic data with properties similar to real data is a fundamental and challenging problem in data science practice. The authors may wish to comment on recent advances in this area that aim to address this challenge without fully specifying a probabilistic model.

**\textcolor{purple}{The reviewer is correct that in certain scenarios (e.g. GEE and GMM), it can be challenging to accurately simulate a dataset that resembles the observed data. However, the primary aim of our paper is not to recover the true data generating process or to find the model that best fits the data. Instead, our goal is to understand which characteristic of the data may lead to the unexpected results - a mismatch between the data and the model. Here, we hold the model fixed and examine the idiosyncrasies in the data that could causes such incompatibility. This contrasts with the common approach of treating data as fixed and focusing on improving the model. Our perspective is motivated by applied studies, where certain modelling strategies are generally accepted for specific types of data. When unexpected outcomes arise, analysts often first investigate whether particular characteristics of the data may cause the results, rather than developing new models. In this context, the analysis validation checks we propose provide a pre-developed "diagnostic checks" to investigate the data. Note that even when the results are as-expected, \textit{we do not assume the model is correct}. For generating synthetic data with similar properties to real data, we agree that this is a challenging area worthy of further investigation. We have added material in the discussion regarding the potential of using recent work in data thinning, data fission, and differential privacy.}**

@@

3. The construction of validation checks appears to be a critical step in applying the proposed methodology. It would be valuable if the authors could offer some guidance or recommendations on the principles for designing these checks. Clarifying this aspect would enhance the practical utility of the approach and help readers apply it more effectively in real-world settings.

**\textcolor{purple}{In the toy example section (Section 3.1), we include some explanations on the construction of validation checks: these are diagnostics analysts may naturally check when they receive an unexpected outcome. For example, when the mean of a vector falls outside an expected range, analysts may check the the five-number summary or plot the variable distribution. The validation checks are those diagnostics framed as binary predicates.}**

4. The independence metric appears to be ill-defined or uninformative for single checks, much like defining total correlation for a single variable. By definition, the metric seems to always take the value one in such cases, as observed in Tables 1 and 3. Including it in the total score may artificially inflate the overall score when compared to checks involving multiple components. Further, in the current examples, the multivariate checks consistently yield independence values close to one. Are there cases where this metric meaningfully deviates from one? Clarifying the interpretation and role of this metric, particularly in the single-check case, would strengthen the methodology.

**\textcolor{purple}{After reviewing the definition of the independence metric, we find it is the previous version is less ideal and we have refined the definition. The amount of overlapping information among checks can be represented by the normalized total correlation, $C' = \frac{C(X_1, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)} = \frac{\sum_{i= 1}^k H(X_i) - H(X_1, X_2, \cdots, X_k)}{\sum_{i= 1}^k H(X_i)}$, where $H(X_i)$ is the entropy of check $X_i$, $C(X_1, \cdots, X_k)$ and $H(X_1, \cdots, X_k)$ are the total correlation and joint entropy of checks $\{X_1, \cdots, X_k\}$, respectively. The amount of independent information is then $\eta = 1 - C' = \frac{ H(X_1, \cdots, X_k)}{\sum_{i = 1}^k}$. The independence metric is now defined as  $\frac{\eta - 1/k}{1 - 1/k}$ to scale $\eta$ to [0, 1], where 0 is attained when all checks are independent and 1 is attained when checks are all identical. A trivial case occurs when there is only one single check, in which case both the joint entropy and the sum of all entropies is the entropy of the check itself, hence giving an independence of 1. These changes have been reflected in Section 4.2 and in the examples.}**

@@

5. The writing is generally clear and easy to follow. However, there are a number of minor syntax errors that could be addressed with more careful proofreading. The use of AI-based writing assistance tools could be helpful in this regard and, in my view, would constitute an ethical and constructive application of such tools.

**\textcolor{purple}{We have run though the spelling check on the article and fixed the spelling mistakes.}**

## Specific Comments

1.  Section 4, paragraph 2: Should the “simulated datasets” be independent of the observed data? If so, this is a big difference from bootstrap samples and a remark would be useful. **\textcolor{purple}{The simulated dataset in our paper are not necessarily independent of the observed data, and we do not rely on bootstrap sampling at any point, although it can be a method to generate the simulated data. We have included a discussion of the principle and different strategies to generate the simulated data after paragraph 2 in Section 4.}**

@@

5.  Section 4.2, paragraph 2: What is the range of total correlation? My understanding is that it is \[0, ∞). Even after standardizing to per observation, the range may still be not \[0, 1) as desired. **\textcolor{purple}{We have re-defined the independence metric (See General comments 4) and the metric is now bounded in [0, 1].}**

6.  Continue: p is undefined; the equation needs to close with a period.  **\textcolor{purple}{We have added the definition of p(.) and close the sentence.}**

7.  Table 2: How did the authors come up with these checks? Readers would appreciate the motivations of these checks. Some of the checks are apparently correlated (e.g., the four checks on mortality-temperature correlation). **\textcolor{purple}{Four mortality-pm10/ mortality-temperature tests are constructed because analysts may not be sure what the best cutoff point for the correlation tests would be. This example illustrates how the method developed can be used to determine the cutoff value for tests.}**

8.  Section 5.1, paragraph 1: Gaussian copula does not allow tail dependence. So by imposing a Gaussian copula, it is assumed that there is no tail dependence in these three variables. As questioned in one of my general comments, this might not be desirable. **\textcolor{purple}{While using a different copula (t- copula or extreme value distribution copula) to better model the tail dependence could produce better simulated data to model temperature and mortality, extreme high temperature does not always correlate with extreme high PM10 in the air pollution literature (hence its role as a confounder can vary). We have maintained the current simulation scheme but added a discussion on this matter in the discussion section.}**

@@

10. Figure 5 caption: The negative sign in front of 0.03 needs to be fixed. **\textcolor{purple}{Fixed.}**

11. Section 5, paragraph 1: The “our second example” appears to be a continuation of the same analysis. **\textcolor{purple}{We use the wording *second example* because we count the toy example as the first example. But, to avoid this confusion, we have changed the wording to *In the following example*.}**

12. Table 4: There are four checks for the mortality-temperature correlation, but only one check reported in this table. **\textcolor{purple}{Addressed in Specific comments 7.}**

13. Section 5: It’d be valuable to see some real improvements from the checks. For example, fit a poor model first and identify the issues from the checks; then fit a strong model to fix the issues and pass the checks. **\textcolor{purple}{We appreciate the reviewer's suggestion, but would like to emphasize that our perspective here is slightly different. The reviewer's comment assumes we always need a better model when an unexpected outcome occurs. However, in practice, it is often likely that there is an issue with the data themselves, warranting at least an initial examination of the data collection or pre-processing aspects, not a revision of the model (although the model may be ultimately updated). Our motiviation in developing the analysis validation checks is primarily to develop a better understanding of the data, not to identify a best-fitting model (This point is also addressed in General comments 2)}**

### 2025-07-22 index.qmd
fix reference

---
title: "Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks"
abstract: "In data analysis, unexpected results often prompt researchers to revisit their procedures to identify potential issues. While some researchers may struggle to identify the root causes, experienced researchers can often quickly diagnose problems by checking a few key assumptions. These checked assumptions, or expectations, are typically informal, difficult to trace, and rarely discussed in publications. In this paper, we introduce the term *analysis validation checks* to formalize and externalize these informal assumptions. We then introduce a procedure to identify a subset of checks that best predict the occurrence of unexpected outcomes, based on simulations of the original data. The checks are evaluated in terms of accuracy, determined by binary classification metrics, and independence, which measures the shared information among checks. We demonstrate this approach with a toy example using step count data and a generalized linear model example examining the effect of particulate matter air pollution on daily mortality. "
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
    attributes:
        corresponding: true
  - name: Roger D. Peng
    email: roger.peng@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
cite-method: natbib
bibliography: references.bib
format:
 tandf-pdf:
   keep-tex: true
   fontsize: 12pt
   linestretch: 1.15
   notebook-links: false
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#remotes::install_github("huizezhang-sherry/adtoolbox")
library(tidyverse)
library(patchwork)
library(MASS)
library(ggh4x)
library(adtoolbox)
library(broom)
library(knitr)
library(rpart)
library(infotheo)
library(patchwork)
library(readr)
library(kableExtra)
library(scales)
library(qqplotr)
pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
  filter(!is.na(pm10)) |>    
  mutate(season = as.factor(season))  
```

@@

$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k)$$

@@

The typical approach to studying the association between PM10 and mortality is to apply a generalized linear model with a Poisson link to relate daily mortality counts to daily measures of PM10. Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between $[0, 0.005]$, after adjusting for daily temperature [@samet2000fine; @welty2005acute]. Note that the relatively simple modeling approach being used here is primarily for demonstration; typically far more sophisticated semi-parametric approaches are used in the literature [@peng2006model].

@@

tbl <- tibble(check = c("Check 1: cor(m, PM10) < -0.03", "Check 2: cor(m, tmp) > -0.35",
                        "Check 3: mortality outlier", "Check 4: PM10 outlier", 
                        "Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4))")) |> 
  bind_cols(df |> dplyr::select(-id)) |> rename(`indep.` = independence)

@@

  > (cor(mortality, PM10) < -0.03) AND (cor(mortality, tmp) > -0.3 OR (mortality outlier AND PM10 outlier))

@@

```{r}
#| label: tbl-all-cities
#| tbl-cap: Results from full analysis of 40 cities from the NMMAPS air pollution and mortality study. In the table, a "T" indicates that a check failed and an "F" indicates that a check passed.

@@

df |>
  mutate(pm10 = format(round(pm10, 4), nsmall = 4),
         cor_m_tmp = format(round(cor_m_tmp, 2), nsmall = 2),
         cor_m_pm10 = format(round(cor_m_pm10, 2), nsmall = 2),
         across(c(cor_m_pm10, cor_m_tmp), ~str_replace(.x, "-", "--"))) |> 
  mutate(across(c(unexpect, .pred, lgl_cor_m_pm10, lgl_cor_m_tmp, mort_outlier, pm10_outlier), 
                ~ifelse(.x, "T", "F"))) |>
  mutate(cor_m_tmp = paste0(cor_m_tmp, " (", lgl_cor_m_tmp, ")"),
         cor_m_pm10 = paste0(cor_m_pm10, " (", lgl_cor_m_pm10, ")")) |> 
  dplyr::select(-lgl_cor_m_tmp, -lgl_cor_m_pm10) |> 
  kable(format = "latex", booktabs = TRUE, escape = F, 
        align = "lrccrrcc",
        col.names = linebreak(c("City", 
                                "PM10\nEstimated\nCoefficient",
                                "Outcome\nObserved\nUnexpected", 
                                "Logic Regression\nPredicted\nUnexpected", 
                                "Correlation:\nMortality-\nTemperature",
                                "Correlation:\nMortality-\nPM10",
                                "Mortality\nOutlier", 
                                "PM10\nOutlier"))) |>
  kableExtra::kable_styling(latex_options="scale_down") 
```

>>>>

---
title: "Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks"
abstract: "In data analysis, unexpected results often prompt researchers to revisit their procedures to identify potential issues. While some researchers may struggle to identify the root causes, experienced researchers can often quickly diagnose problems by checking a few key assumptions. These checked assumptions, or expectations, are typically informal, difficult to trace, and rarely discussed in publications. In this paper, we introduce the term *analysis validation checks* to formalize and externalize these informal assumptions. We then introduce a procedure to identify a subset of checks that best predict the occurrence of unexpected outcomes, based on simulations of the original data. The checks are evaluated in terms of accuracy, determined by binary classification metrics, and independence, which measures the shared information among checks. We demonstrate this approach with a toy example using step count data and a generalized linear model example examining the effect of particulate matter air pollution on daily mortality. "
author:
  - name: H. Sherry Zhang
    email: huize.zhang@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
    attributes:
        corresponding: true
  - name: Roger D. Peng
    email: roger.peng@austin.utexas.edu
    affiliations: 
        - id: 1
          name: University of Texas at Austin
          department: Department of Statistics and Data Sciences
          city: Texas
          country: United States
          postal-code: 78705
cite-method: natbib
bibliography: references.bib
format:
 tandf-pdf:
   keep-tex: true
   fontsize: 12pt
   linestretch: 1.15
   notebook-links: false
   keywords: diagnostic, logic regression, data analysis assumptions
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
#remotes::install_github("huizezhang-sherry/adtoolbox")
library(tidyverse)
library(patchwork)
library(MASS)
library(ggh4x)
library(adtoolbox)
library(broom)
library(knitr)
library(rpart)
library(infotheo)
library(patchwork)
library(readr)
library(kableExtra)
library(scales)
library(qqplotr)
pm10 <- read_csv(here::here("data", "pm10.csv")) |> 
  filter(!is.na(pm10)) |>    
  mutate(season = as.factor(season))  
```

@@

$$E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k),$$

@@

The typical approach to studying the association between PM10 and mortality is to apply a generalized linear model (Poisson family with a log link) to relate daily mortality counts to daily measures of PM10. Based on previous work and the range of effect sizes published in the literature, an analyst might expect the coefficient for PM10 in this GLM to lie between $[0, 0.005]$, after adjusting for daily temperature [@samet2000fine; @welty2005acute]. Note that the relatively simple modeling approach being used here is primarily for demonstration; typically far more sophisticated semi-parametric approaches are used in the literature [@peng2006model].

@@

tbl <- tibble(check = c("Check 1: cor(m, PM10) < -- 0.03", "Check 2: cor(m, tmp) > -- 0.35",
                        "Check 3: mortality outlier", "Check 4: PM10 outlier", 
                        "Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4))")) |> 
  bind_cols(df |> dplyr::select(-id)) |> rename(`indep.` = independence)

@@

  > (cor(mortality, PM10) < -- 0.03) AND (cor(mortality, tmp) > -- 0.3 OR (mortality outlier AND PM10 outlier))

@@

```{r message=FALSE}
#| label: tbl-all-cities
#| tbl-cap: Results from full analysis of 40 cities from the NMMAPS air pollution and mortality study. In the table, a "T" indicates that a check failed and an "F" indicates that a check passed.

@@

df |>
  mutate(pm10 = format(round(pm10, 4), nsmall = 4),
         cor_m_tmp = format(round(cor_m_tmp, 2), nsmall = 2),
         cor_m_pm10 = format(round(cor_m_pm10, 2), nsmall = 2),
         across(c(pm10, cor_m_pm10, cor_m_tmp), ~str_replace(.x, "-", "--"))) |> 
  mutate(across(c(unexpect, .pred, lgl_cor_m_pm10, lgl_cor_m_tmp, mort_outlier, pm10_outlier), 
                ~ifelse(.x, "T", "F"))) |>
  mutate(cor_m_tmp = paste0(cor_m_tmp, " (", lgl_cor_m_tmp, ")"),
         cor_m_pm10 = paste0(cor_m_pm10, " (", lgl_cor_m_pm10, ")")) |> 
  dplyr::select(-lgl_cor_m_tmp, -lgl_cor_m_pm10) |> 
  kable(format = "latex", booktabs = TRUE, escape = F, 
        align = "lrccrrcc",
        col.names = linebreak(c("City", 
                                "PM10\nEstimated\nCoefficient",
                                "Outcome\nObserved\nUnexpected", 
                                "Logic Regression\nPredicted\nUnexpected", 
                                "Correlation:\nMortality-\nTemperature",
                                "Correlation:\nMortality-\nPM10",
                                "Mortality\nOutlier", 
                                "PM10\nOutlier"))) |>
  kableExtra::kable_styling(latex_options="scale_down") 
```
